Fast and adaptive cointegration based model for forecasting high frequency financial time series

### 2017-01-20 articles-conferences/iscef2016/body.tex
adding conclusions
correccion Jon

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that some sets
of variables cannot wander too far from each other. Cointegration means that one
or more linear combinations of these variables are stationary even though
individually they are not. Some models, such as the Vector Error Correction
(VECM), see \cite{engle87}, take advantage of this property and describe the
joint behaviour of several cointegrated variables. VECM introduces this long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}).

Cointegration relationship can be found in low and high frequency data of two
or more assets. While cointegration in low frequency data is motivated by a
long-run equilibrium relationship between economic forces, cointegration in
high frequency data has its foundation in statistical arbitrage which is very
helpful to detect mean-reverting trades. Cointegration at different data
frequencies may differ. \cite{rittler2012} shows how cointegration appears when
the data frequency is increased.

The use of VECM with high frequency data is mainly limited by computational 
expensive routines. Firstly, VECM parameters are obtained using the ordinary least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves many calculations, the
parameter estimation is computationally expensive when the number of lagged
values and data increases. Secondly, obtaining cointegration vectors is also an
expensive routine because the Johansen method, see \cite{johansen1995}, is required, which is of order $O(n^3)$. \cite{chen2003} addresses the advantage of distributed
processing over conventional rolling window processing. Therefore, our aims were to study if a parallel version of VECM can be used with high frequency stream data.

Our approach was to determine adaptively the number of observations and lags of
VECM which maximise cointegration relations in the past in a distributed environment. 
We called Adaptively Vector Error Correction (AVECM) to our proposal. AVECM
parallelise this search of parameters in order to update them before new data
arrives. Model effectiveness is focused on out-of-sample forecast rather than
in-sample fitting. This criterion allows AVECM prediction capability to be
expressed rather than just explaining data history.  The forecast capability of
our method was measured using MSE and the Theil's $U$-statistic, see
\cite{theil1966}, widely used in economic forecast. Tests were run using four
currency rates: Euro (EUR) to United States Dollar (USD) (EURUSD), British
Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF) and USD to
Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

@@

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
long-run relationships between the time series  might change due to several
economic factors that can lead to structural breaks in the cointegration
relationship.  In order to show that the number of cointegration vectors depends
on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
used a grid search.  We defined a grid of possible values for $L$ and $p$. $L$
goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a step size of
4 hours and $p$ throughout $[1,5]$ with step size of 1. The first experiment
consisted in determining the number of cointegration vectors for all
combinations of $L$ and $p$.  We used four forex rates: EURUSD, GBPUSD, USDCHF
and USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of
August 2014, when the New York and London financial markets opened.

Figure \ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$
hours  and $p=1$ . This procedure was carried out by a sliding window of
historical data moving 1000 times.

@@

@@

The goal of our next experiment was to find a relation between this ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[700,1500]$ data points with step size 100 and $p$ took values between
$[1,5]$ with step size 1.

\begin{figure}[ht!]
  \centering
  %\includegraphics[width=0.9\textwidth]{Fig2}
   \includegraphics[width=0.9\textwidth]{img/Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Best windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 80\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relation between MSE and $PC$ and $L$.  We
found that better cointegration percentage leads to better performance accuracy
in terms of reducing MSE.

@@

\section{Experimental results}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
\cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign
Exchange marketplace.

@@

The parallel implementation allowed the execution times to be reduced more than
9 times and therefore a response time was obtain before 10 seconds. Since we
used 10-second frequencys we can say that our proposal is suitable for use in
an online context for real applications because response times were less than
this frequency.

>>>>

\section{Introduction}
\label{sec:introduction}
In finance, it is common to find variables with long-run equilibrium
relationships. This is termed cointegration and reflects the idea that some sets
of variables cannot wander too far from each other. Cointegration means that one
or more linear combinations of these variables are stationary even though
individually they are not. Some models, such as the Vector Error Correction
(VECM), see \cite{engle87}, take advantage of this property and describe the
joint behaviour of several cointegrated variables. VECM introduces this long-run
relationship among a set of cointegrated time series as an error correction
term. These time series must be integrated of order 1, denoted I(1), i.e. they
become stationary at their first differences. In finance, I(1) time series are
very common and to introduce cointegration restrictions in models often improves
forecasting, see \cite{duy1998}. Therefore, VECM has been widely adopted in
financial applications, among others: \cite{mukherjee1995}, \cite{seong2013},
\cite{maysami2000} and \cite{arestis2001}. VECM has also been used in pair
trading, see \cite{herlemont2003}, or models with more than two variables, see
for example \cite{mukherjee1995} and \cite{engle2004}.

Cointegration relationships can be found in low and high frequency data of two or
more assets. While cointegration in low frequency data is motivated by a
long-run equilibrium relationship between economic forces, cointegration in high
frequency data has its foundation in statistical arbitrage which is very helpful
to detect mean-reverting trades. Information about cointegrated assets in high
frequency data could be used as an input for high frequency trading strategies,
using it as a signal to capture small profits in short term trades, see
\cite{miao2014}. \cite{zhou2001} addressed the benefits of using higher
frequency data to analize cointegration. \cite{rittler2012} also found that
cointegration may differ with different data frequencies and could appear
whith increased data frequency.

The use of VECM with high frequency data is mainly limited by computationally
expensive routines. Firstly, VECM parameters are obtained using the ordinary
least squares (OLS) method, developed by \cite{golub1980}. Since OLS involves
many calculations, the parameter estimation is computationally expensive when
the number of lagged values and data increases. Secondly, obtaining
cointegration vectors is also an expensive routine because the Johansen method 
is required, which is of order $O(n^3)$ (\cite{johansen1995}).
\cite{chen2003} addressed the advantage of distributed processing over
conventional rolling window processing. Therefore, our aim was to study if a
parallel version of VECM can be used with high frequency stream data.

Our approach was to determine, adaptively, the number of observations and lags of
VECM which maximise cointegration relations in the past in a distributed
environment. We called our proposal Adaptive Vector Error Correction (AVECM). 
AVECM parallelises this search of parameters in order to update them
before new data arrives. Model effectiveness is focused on out-of-sample
forecast rather than in-sample fitting. This criterion allows AVECM prediction
capability to be expressed rather than just explaining data history.  The
forecast capability of our method was measured using MSE and the Theil's
$U$-statistic, see \cite{theil1966}, widely used in economic forecast. Tests
were run using four currency rates: Euro (EUR) to United States Dollar (USD)
(EURUSD), British Pound (GBP) to USD (GBPUSD), USD to Swiss Franc (CHF) (USDCHF)
and USD to Japanese Yen (JPY) (USDJPY) with a 10-second frequency.

@@

Cointegration vectors can be found applying the Johansen method which uses a
sample of the last historical data. However, VECM assumes cointegration vectors
do not change in time.  In fact, \cite{gregoryETal1996} addresses that the
long-run relationships between the time series  might change due to several
economic factors that can lead to structural breaks in the cointegration
relationship.  In order to show that the number of cointegration vectors depends
on the amount $L$ of historical data and the number of lags $p$ in the VECM, we
used a grid search.  We arbitrarily defined a grid of possible values for $L$
and $p$. $L$ goes throughout $[2,14]$ hours (1 $hour = 360$ data points) with a
step size of 4 hours and $p$ throughout $[1,5]$ with a step size of 1. The idea
was to show the variability of the number of cointegration vectors when we changed
these two parameters. We used four forex rates: EURUSD, GBPUSD, USDCHF and
USDJPY with 10-second frequency. Data started at 13:00 GMT of the 13th of August
2014, when the New York and London financial markets opened.

@@

Figure \ref{fig:hists} shows the distribution of the number of cointegration
vectors given by the Johansen method for different values of $L=[2,6,10,14]$
hours and $p=1$ . This procedure was carried out by a sliding window of
historical data moving 1000 times. We observed that the distribution of
cointegration vectors changed with different values of $L$. When
$L=2$ hours, there was no cointegration in more than 60\% of the iterations.
Cointegration increased when $L=6$ hours and was maximum when $L=10$ hours where
one cointegration vectors was found for all 1000 iterations. The ocurrence of
cointegration started to decrease at $L=14$ hours.

@@

The goal of our next experiment was to find a relationship between the ratio $PC$
and the performance measure MSE (see equation \ref{eq:MSE}). $L$ was defined
between $[2,14]$ hours, that corresponded to $[720, 5040]$ data points, and $p$
took values between $[1,5]$.

\begin{figure}[ht!]
  \centering
  %\includegraphics[width=0.9\textwidth]{Fig2}
   \includegraphics[width=0.9\textwidth]{img/Fig2}
  \caption{MSE versus the percentage of cointegration considering 1000
  iterations. Optimum windows size $L$ found was 1060. Below MSE versus $L$ shows a
  rapidly decreasing behaviour, founding minimum at PC = 73\% }
  \label{fig:cointvsmse}
\end{figure}

Figure~\ref{fig:cointvsmse} shows the relationship between MSE and $PC$ and $L$. We
found that higher cointegration percentage leads to improved performance accuracy
in terms of lower MSE. Also, increasing the size of the sliding window $L$
doesn't necessarily help to reduce MSE.

@@

\section{Experimental results}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
All the experiments and AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. We chose the most traded rates related
to USD so they were likely to be cointegrated. 
This data was collected from
\cite{Dukascopy2014}, a free database which gives access to the Swiss Foreign
Exchange marketplace.

@@

The parallel implementation allowed the execution times to be reduced more than
9 times and therefore a response time was obtain before 10 seconds. Since we
used 10-second frequencys we can say that our proposal is suitable for use in an
online context for real applications because response times were less than this
frequency. Cointegration information can now easily be used as an integration tool
to detect arbitrage opportunities or risk control.
